Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-03-17 paper.tex
section 6

\textbf{First Version of Evaluation Function.}
Since our first version of @aeval@ just implements the nondeterministic
evaluation, and given the defined @nd@, the evaluation function @aeval@ can be easily
defined. For the case that @e@ is a @let@ expression,
the overall idea is to perform a two step depth-first evaluation. \todo{refine}

@@

The case class @Cache@ is defined as follow:

@@

The @Cache@ contains two maps @in@ and @out@ which are both mapping from 
configurations to sets of @VS@.

When we enter to function @aeval@

\textbf{Cache-Passing Style.}

The definition of Ans is changed to a case class that contains a set of @VS@ and
a cache.
\begin{lstlisting}
case class Ans(vss: Set[VS], cache: Cache)
\end{lstlisting}

\begin{lstlisting}
def aeval(e: Expr, env: Env, store: BStore, time: Time, cache: Cache, continue: Cont): Ans = {
  val config = Config(e, env, store, time)
  if (cache.outContains(config)) return continue(Ans(cache.outGet(config), cache))

  val new_cache = cache.outUpdate(config, cache.inGet(config))
  val new_time = (e::time).take(k)

  e match {
    case Let(x, App(f, ae), e) =>
      val closures = atomicEval(f, env, store).asInstanceOf[Set[Clos]]
      nd[Clos, Ans](closures, Ans(Set[VS](), new_cache), { case (clos, Ans(acc, cache), closnd) =>
        val Clos(Lam(v, body), c_env) = clos
        val baddr = allocBind(v, new_time)
        val new_env = c_env + (v -> baddr)
        val new_store = store.update(baddr, atomicEval(ae, env, store))
        aeval(body, new_env, new_store, new_time, cache, { case Ans(bodyvss, bodycache) =>
          nd[VS, Ans](bodyvss, Ans(Set[VS](), bodycache), { case (vs, Ans(acc_vss, cache), bdnd) =>
            val VS(vals, time, store) = vs
            val baddr = allocBind(x, time)
            val new_env = env + (x -> baddr)
            val new_store = store.update(baddr, vals)
            aeval(e, new_env, new_store, time, cache, { case Ans(evss, ecache) => bdnd(Ans(acc_vss ++ evss, ecache)) })
          },
          { case Ans(evss, cache) => closnd(Ans(evss ++ acc, cache)) })
        })
      },
      { case Ans(vss, cache) => continue(Ans(vss, cache.outUpdate(config, vss))) })

    case ae if isAtomic(ae) =>
      val vs = Set(VS(atomicEval(ae, env, store), new_time, store))
      continue(Ans(vs, new_cache.outUpdate(config, vs)))
  }
}
\end{lstlisting}

>>>>

\textbf{Evaluation without Caching.}
Since our first version of @aeval@ just implements the nondeterministic
evaluation, and given the defined @nd@, the evaluation function @aeval@ can be easily
defined. For the case that @e@ is a @let@ expression,
the overall idea is to perform a two step depth-first evaluation. \todo{refine}

@@

The case class @Cache@ and its operations are defined as follow:

@@

The @Cache@ contains two maps @in@ and @out@ which are both mapping from 
configurations to sets of @VS@. The @in@ cache contains mappings from 
last iteration of evaluation, and @out@ cache contains mappings after this
iteration of evaluation. Once we have evaluated a term to some values, 
we update the @out@ cache. In the next time of iteration, we will use @out@
from last iteration as @in@ cache for this iteration.

\textbf{Cache-Passing Style.}
Now we can modify out interpreter to cache-passing style. 
First the return type @Ans@ is changed to a case class that contains a set of @VS@ and
a cache.

\begin{lstlisting}
case class Ans(vss: Set[VS], cache: Cache)
\end{lstlisting}

The skeleton of function @aeval@ largely remain the same. 
Every place has a value of @Ans@ is also replaced as direct pattern matching so 
that we can use the fields in @Ans@ immediately.
The cache is used in a monotonic way: each function call to @aeval@ or @nd@ is
passed with the latest cache from the most recent continuation's result.

When we enter to function @aeval@, we firstly lookup in the @out@ cache, 
if there are some values for current configuration, then we use immediately.
This corresponds to small-step AAM with a worklist when we have seen a state 
then we can simply discard the state and continue working the rest of worklist @todo@.

Otherwise, if we not hit the @out@ cache, we retrieve the values from @in@ cache, and update @out@ cache
with values from @in@ cache.  The retrieving from @in@ cache may return an empty set of values if 
there is no such mapping for the configuration from the previous iteration.
It actually first conservatively assumes that all computation can be diverged,
if not, we will update its values in the mapping after evaluation.
In terms of partial orders and lattices, it is also the case we start 
the Kleene iteration from the bottom of the lattice.

\begin{lstlisting}
def aeval(e: Expr, env: Env, store: BStore, time: Time, cache: Cache, continue: Cont): Ans = {
  val config = Config(e, env, store, time)
  /* lookups the out cache, uses the values if contains this config */
  if (cache.outContains(config)) return continue(Ans(cache.outGet(config), cache))
  /* uses the values from in cache */
  val new_cache = cache.outUpdate(config, cache.inGet(config))
  val new_time = (e::time).take(k)
  e match {
    case Let(x, App(f, ae), e) =>
      val closures = atomicEval(f, env, store).asInstanceOf[Set[Clos]]
      nd[Clos, Ans](closures, Ans(Set[VS](), new_cache), { case (clos, Ans(acc, cache), closnd) =>
        ......
        aeval(body, new_env, new_store, new_time, cache, { case Ans(bodyvss, bodycache) =>
          nd[VS, Ans](bodyvss, Ans(Set[VS](), bodycache), { case (vs, Ans(acc_vss, cache), bdnd) =>
            ......
            aeval(e, new_env, new_store, time, cache, { case Ans(evss, ecache) => 
              bdnd(Ans(acc_vss ++ evss, ecache)) 
            })
          },
          { case Ans(evss, cache) => closnd(Ans(evss ++ acc, cache)) })
        })
      },
      /* updates the out cache after evaluation */
      { case Ans(vss, cache) => continue(Ans(vss, cache.outUpdate(config, vss))) })

    case ae if isAtomic(ae) =>
      val vs = Set(VS(atomicEval(ae, env, store), new_time, store))
      continue(Ans(vs, new_cache.outUpdate(config, vs))) /* updates the out cache after evaluation */
  }
}
\end{lstlisting}

After done the evaluation, we obtain a set of @VS@ and need to update the @out@ cache.
In the first case of pattern matching, it happens inside of the last continuation of 
outer @nd@. 
For the second case, we update the @out@ cache before calling the continuation @continue@.
